Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-06-04 WSSSPE1_report/WSSSPE1_summary_report.tex
edits in project years and style for showing them

In this section, we discuss some of the software projects as case
studies to better understand the points discussed during the workshop
and described in the previous sections, and to find how 
they are affected by sustainability issues in
practice. 
%Additionally, we discuss issues surrounding people,
%especially the career paths of original developers and investigate to
%answer the question: \emph{are the original developers still around?}
%
In terms of age, most of the software discussed here were publicly launched 10
or more years ago. To get a sense of their age, we note release year in the
parenthesis at first mention of the software where available.

We classify the software projects discussed in the workshop in two
broad categories. First, the \emph{utility} software, comprising of
general purpose software. Utility software is often used as enabler
or facilitator for the development of other tools and techniques
to carry out scientific work. This includes the software developed to
efficiently utilize new research infrastructures. Second, the
\emph{scientific} software, comprising the software that was
originally developed with an aim to solve a specific scientific
problem.
%
% Why are we classifying software like above?
This classification is motivated by our argument that the two kinds of software
projects wildly vary in factors such as scope, purpose and usage. The
development and management of each kind is significantly different.
Consequently, the sustainability challenges faced by them differ and must be
treated separately. For instance, the challenges faced by a gateway software
development project such as \toolname{CIPRES (r2007)} or visualization software
products such as \toolname{VisIT (r2001)} or \toolname{ParaView (r2002)} are
distinct to a niche software for \textit{ab initio} modeling and simulation
such as \toolname{VASP (r1992)} or \toolname{Quantum Espresso}. %\note{cite SW}

\subsection{Utility Software}
Software developed with a potentially wider audience and general purpose usage
in mind is utility software. Utility software typically does not address
fundamental research problems for a given scientific domain. Examples are
collaborative development frameworks such as \toolname{GitHub (r2008)} and
\toolname{Bitbucket (r2008)}, distributed workflow and generic computing
frameworks such as \toolname{Galaxy (r2006)}, \toolname{HUBzero},
\toolname{SimGrid}, \toolname{Swift (r2007)}, \toolname{Globus (r2000)} and
\toolname{VisTrails}, and visualization frameworks such as \toolname{VTK},
\toolname{VisIT}, and \toolname{ParaView}.

@@

One such example is the \toolname{Galaxy} project~\cite{Galaxy}. It follows
agile software development practices and implements standards such as
test-driven development and socialized bug managing practices via
\toolname{trello}. \toolname{Galaxy} \emph{histories} and \emph{toolshed} offer
easy community sharing of data and tools further promoting a collaborative
environment. The project closely follows the guidelines described in Carver and
Thiruvathukal~\cite{Carver_WSSSPE} and many from Prli\'{c} and
Procter~\cite{Prlic_WSSSPE}. Many utility software projects are often developed
aiming better utilizing a particular, new infrastructure and architecture,
e.g., \toolname{MVAPICH (r2002)}, \toolname{VisIT}, \toolname{ParaView}. Similarly, to
leverage the power of emerging architectures such as accelerators, new code and
libraries are required. The experience of one such effort as described in
Ferenbaugh~\cite{Ferenbaugh_WSSSPE} which met with a limited success but
nonetheless with many invaluable lessons were learned about influential
cultural and technical aspects in sustainable software development practices.

@@

Some examples of such software discussed as part of the workshop are,
\toolname{DUNE (r2008)}, \toolname{R/qtl (r2002)}, \toolname{Kitware (r1998)},
\toolname{PETSc (r1995)}, \toolname{MINRES-QLP}, \toolname{FASTMath}.  Most of
the above efforts are focused on one scientific or applied
mathematics domain. However, sometimes such projects grow beyond the
initial vision of developers. One such example is \toolname{Kitware},
which while being a software product specializing in the scientific
process, has a core focus of developing communities around software
processes. Another instance of this process is the development of the
\toolname{CMake} build utility, which started out as a building tool
for \toolname{ITK} but grew to become a generic build utility for C++
projects. Similarly, \toolname{PETSc} is growing towards becoming a
general purpose utility system usable for solving a variety of
scientific problems.

>>>>

In this section, we discuss some of the software projects as case
studies to better understand the points discussed during the workshop
and described in the previous sections, and to find how 
they are affected by sustainability issues in
practice. 
%Additionally, we discuss issues surrounding people,
%especially the career paths of original developers and investigate to
%answer the question: \emph{are the original developers still around?}
%
Most of the software projects discussed here were publicly launched 10
or more years ago. We generally note the original release (o.r.) year of each project in 
parenthesis in its first mention.

We classify the software projects discussed in the workshop in two
broad categories. First, the \emph{utility} software, comprising of
general purpose software. Utility software is often used as enabler
or facilitator for the development of other tools and techniques
to carry out scientific work. This includes the software developed to
efficiently utilize new research infrastructures. Second, the
\emph{scientific} software, comprising the software that was
originally developed with an aim to solve a specific scientific
problem.
%
% Why are we classifying software like above?
This classification is motivated by our argument that the two kinds of software
projects wildly vary in factors such as scope, purpose and usage. The
development and management of each kind is significantly different.
Consequently, the sustainability challenges faced by them differ and must be
treated separately. For instance, the challenges faced by a gateway software
development project such as \toolname{CIPRES} (o.r. 2007) or visualization software
products such as \toolname{VisIT (o.r. 2001)} or \toolname{ParaView} (o.r.  2002) are
distinct to a niche software for \textit{ab initio} modeling and simulation
such as \toolname{VASP} (o.r. 1992) or \toolname{Quantum Espresso} (o.r. 2001). %\note{cite SW}

\subsection{Utility Software}
Software developed with a potentially wider audience and general purpose usage
in mind is utility software. Utility software typically does not address
fundamental research problems for a given scientific domain. Examples are
collaborative development frameworks such as \toolname{GitHub} (o.r. 2008) and
\toolname{Bitbucket} (o.r.  2008), distributed workflow and generic computing
frameworks such as \toolname{Galaxy} (o.r. 2006), \toolname{HUBzero} (o.r. 2010),
\toolname{SimGrid} (o.r. 2001), \toolname{Swift} (o.r. 2007), \toolname{Globus} (o.r. 2000) and
\toolname{VisTrails} (o.r. 2007), and visualization frameworks such as \toolname{VTK},
\toolname{VisIT}, and \toolname{ParaView}.

@@

One such example is the \toolname{Galaxy} project~\cite{Galaxy}. It follows
agile software development practices and implements standards such as
test-driven development and socialized bug managing practices via
\toolname{trello}. \toolname{Galaxy} \emph{histories} and \emph{toolshed} offer
easy community sharing of data and tools further promoting a collaborative
environment. The project closely follows the guidelines described in Carver and
Thiruvathukal~\cite{Carver_WSSSPE} and many from Prli\'{c} and
Procter~\cite{Prlic_WSSSPE}. Many utility software projects are often developed
aiming better utilizing a particular, new infrastructure and architecture,
e.g., \toolname{MVAPICH} (original release 2002), \toolname{VisIT}, \toolname{ParaView}. Similarly, to
leverage the power of emerging architectures such as accelerators, new code and
libraries are required. The experience of one such effort as described in
Ferenbaugh~\cite{Ferenbaugh_WSSSPE} which met with a limited success but
nonetheless with many invaluable lessons were learned about influential
cultural and technical aspects in sustainable software development practices.

@@

Some examples of such software discussed as part of the workshop are,
\toolname{DUNE (o.r. 2008)}, \toolname{R/qtl} (o.r. 2002), \toolname{Kitware} (o.r. 1998),
\toolname{PETSc} (o.r. 1995), \toolname{MINRES-QLP} (o.r. 2007), \toolname{FASTMath} (o.r. \textcolor{red}{TBD - Is FASTMath really a software project, or a collection of different packages?}).  Most of
the above efforts are focused on one scientific or applied
mathematics domain. However, sometimes such projects grow beyond the
initial vision of developers. One such example is \toolname{Kitware},
which while being a software product specializing in the scientific
process, has a core focus of developing communities around software
processes. Another instance of this process is the development of the
\toolname{CMake} build utility, which started out as a building tool
for \toolname{ITK} but grew to become a generic build utility for C++
projects. Similarly, \toolname{PETSc} is growing towards becoming a
general purpose utility system usable for solving a variety of
scientific problems.
